The effects of symbols and information in government communications on citizens  trust: Evidence from a survey experiment

### 2018-05-21 paper.Rmd
Add files via upload

---
title: 'The effects of symbols and information in government communication on citizens  trust in policy: Evidence from a survey experiment'
author: '**Saar Alon-Barkat**, PhD candidate; the Hebrew University of Jerusalem;   saar.barkat@mail.huji.ac.il'
date: " "
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
  word_document: default
link-citations: yes
bibliography: phd_paper_2.bib
urlcolor: blue
---

@@

To be presented at the 2018 Transatlantic Dialogue (TAD 14) at Workshop 1: *Behavioral Public Administration and Psychological perspectives*.

@@

*Effective public communications have been proposed as a remedy for citizens' distrust in government. The studies that explored the effectiveness of government communications have focused mainly on the extent to which they provide substantive information about bureaucracies' performance and policies. Recently, a few studies pointed to the effect of symbolic elements, entangled in these communications (e.g. logos, images and celebrities). Still, we lack theoretical understanding of the different mechanisms underlying citizens' responses to information and symbolic elements, and accordingly - the interaction between them. Building on the Elaboration Likelihood Model (ELM) from social psychology, I hypothesize that symbolic elements have a greater effect on citizens' trust the less citizens perceive the issue as personally relevant. Also, I hypothesize that symbolic elements are likely to lead citizens to dedicate less attention to the substantive information in the communication, and accordingly, to increase their trust in logically-unpersuasive policy plans. I put these hypotheses to test through a survey experiment, where I examine Israeli citizens' trust in the Environment Protection Ministry's policies for reducing the air pollution in a specific area (the Haifa-Bay) and reducing domestic waste. The policies are presented to subjects under different conditions of substantive information (logically persuasive vs. unpersuasive policies) and symbolic elements (logo, color, and celebrities endorsers). I operationalize personal relevance by comparing between citizens from the polluted Haifa-Bay area and others. The results do not support the hypothesis that personal relevance attenuates the effect of symbols. Rather, they suggest that symbolic elements increase citizens' trust in ill-perceived policies, by distracting them from scrutinizing their content.*

@@

I further postulated, based on ELM, that the effect of symbols is attenuated the more the communication is personally relevant for citizens' lives. However, this latter expectation was not supported by the results. Residents of the polluted Haifa-Bay, who perceive EPM policy plan for reducing air-pollution in that area as highly relevant for their lives, were affected by the symbols not less than those who reside elsewhere. If anything, the results suggest that the symbols had a greater effect on Haifa-bay subjects. How can we reconcile these unexpected findings with the proposed theory? Given that the moderating role of personal relevance on peripheral cues was substantiated by numerous studies in social psychology, I would not argue that we should abandon this postulation based on this one study. One plausible explanation is that although Haifa-bay residents perceived the air-pollution policy as more relevant, their motivation to elaborate was not high enough to cause them to avoid peripheral cues. Certainly, the air-pollution in their area concerns them a great deal, as they point out themselves. However this issue represents a *continuous* concern that they may have got used to over the time, as opposed a new issue that emerged recently. If this explanation holds, this entails that citizens may resist the emotional effect of symbols only in rare circumbstances, which are charecterized by very high levels of personal relevance. These latter findings are also consistent with the findings of @alonbarkat_2017. In that study, the scholars assessed citizens' trust in the government electricity company, and its shaping by its familiar symbol, given citizens' recent personal experience of prolonged blackouts. Whilst their study was not purposly designed to test the interaction between symbols and personal relevance, one can assume that those who personally experienced blackouts perceived electricity service as more personally relevant. Still, they found that citizens who experienced blackouts were influenced by the symbol, as much as those who did not.

Certainly, we must be cautious when drawing general conclusions from our findings, for several particular reasons. First, this study focuses, empirically, on one public organization, which invested greatly in symbolic communications, most prominently, the use of high-profile celebrity comedians in its public campaigns. Future studies may further examine the proposed theory by focusing on additional cases of public organizations and policy areas. Moreover, in the symbols treatment group, I used a bundle of symbols, without dissentangling their distinct effects. Future studies may compare the effectiveness of different types of symbolic elements. Finally, this study focuses on communication about policy plans, and therefore we must be cautious about generalizing from these findings to other types of communications. Specifically, future studies may seek to test my theory with regard to communications that include evidence about performance, and examine the interaction bewteen symbols and information on good/bad outcomes.

Keeping these limitations in mind, this study has several important implications for public administration research. Most importantly, my findings highlight the significat role of branding and symbolic communication in shaping citizens' attitudes about government. Citizens' responses to symbolic elements in communications thus far received relatively little attention in Public administration literature, compared with the abounded research on their responses to information. Yet, this study, and others, demonstrated that symbols can be fairly effective in increasing citizens' trust, and mitigating their negative attitudes [@alonbarkat_2017; @karens_2016; @teodoro_2018]. Furthermore, my findings suggest that symbols decrease citizens' "rational", systematic processing of substantive information. This is good news as far as we are concerned mainly about citizens' *undervaluation* of government performance, and the negative biases that undermine the effectiveness of positive information.[^footnote_biases_ELM]. On the other hand, the emotive effect of government symbols on public opinion is highly problematic, insofar as it challenges the democratic ideal that citizens form their opinions about their government and its actions on the basis of rational thinking [@dahl_1973; @soroka_2010]. Moreover, it becomes especially problematic when symbols are used by poorly performing organizations,  or organizations that lack professional expertiese. It those cases, symbolic elements can mitigate justified public criticism of government failures. Organizations can use them deliberately to distort public opinion via propaganda, or - more likely - without being aware of the tention between the symbols and their substantive message. Ultimately, by preserving citizens' trust in them and escaping from public blame, organizations can nurture their long-term positive reputation, which provides them political power and attenuates elected politicians' incentives to monitor and oversight them [@busuioc_2016; @bertelli_2018]. Hence, whereas government communication with citizens is normally perceived as a tool for enhancing democratic responsiveness and accountability, organizations can also use it as a means to undermine them.

@@

This study also contributes to existing behavioral Public Administration resarch by intoroducing ELM and applying it as a comprehensive theoretical framework for studying citizens' attitudes formation, and citizens' responses to government communications. ELM has various postulations and implications, but most importantly, it stresses that people respond differently to different communications, depending on people's motivation and ability to think about the subject, and on the quality of arguments and the peripheral cues in the communication. In this study, I applyed ELM to distinguish bewteen the peripheral processing of symbolic elements versus the central processing of substantive information, and to exlore their moderation by citizens' motivation to elaborate on the message. Future studies may apply ELM and further explore the effects of other peripheral cues in government communications, and of other relevant motivation and ability factors, to enhance our understanding of the conditions under which government communications are likely to be more/less effective.

Finally, ELM has important methodological implications regarding the external validity of survey experiment studies of citizens' responses to government communications. The participants of these studies, including this particular one, have relatively high motivation to elaborate on the information that scholars presented them, because they may assume that they will be asked about it afterwards. In case they receive payment or other compensation for their participation, this further increases their motivation. By comparison, in the "real world", citizens are far less motivated to invest in thinking about government messages, and they probably also have lower ability to do so, given the high competition on their time and attention. Accordingly, their responses to performance information will less affected by asymetrical processing of information, because most of them are not likely to process this information in the first place. Instead, most of them will rely on simple cues and heuristic shortcuts. I would expect, tentatively, that these differences exacerbate biases due to partisan and political ideology [@james_2017a; @baekgaard_2016] and implicit attitudes about the public-sector [@marvel_2015; @marvel_2015a; @hvidman_2016], while attenuating the role of negativity bias [@nielsen_2016; @olsen_2017]. Future studies may empirically explore these interactions. Regardless, scholars should also take to account this methodological issue when designing survey experiments, and attempt to simulate real world conditions, as much as possible. For instance, they can add filler sections in the survey before the real manipulation to attenuate participants' motivation and ability to elaborate. In addition, because the graphic design of communications also affects motivation and ability to elborate, as shown here, scholars should simulate the design of real communications.

>>>>

---
title: 'The effects of symbols and information in government communications on citizens  trust: Evidence from a survey experiment'
author: '**Saar Alon-Barkat**, PhD candidate; the Hebrew University of Jerusalem;   saar.barkat@mail.huji.ac.il'
date: " "
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
  word_document: default
link-citations: yes
bibliography: phd_paper_2.bib
urlcolor: blue
---

@@

Paper presented at the 2018 Transatlantic Dialogue (TAD 14) at Workshop 1: *Behavioral Public Administration and Psychological perspectives*. June 2018, Bucharest, Romania.

@@

*Effective public communications have been proposed as a remedy for citizens' distrust in government. The studies that explored the effectiveness of government communications have focused mainly on the extent to which they provide substantive information about bureaucracies' performance and policies. Recently, a few studies pointed to the effect of symbolic elements, entangled in these communications (e.g. logos, images and celebrities). Still, we lack theoretical understanding of the different mechanisms underlying citizens' responses to information and symbolic elements, and accordingly - the interaction between them. Building on the Elaboration Likelihood Model (ELM) from social psychology, I hypothesize that symbolic elements are likely to lead citizens to dedicate less attention to the substantive information in the communication, and accordingly, to increase their trust in logically-unpersuasive policy plans. I also hypothesize that symbolic elements have a greater effect on citizens' trust the less citizens perceive the issue as personally relevant. To test these hypotheses, I conducted a survey experiment, focusing on communications of the Israeli Environment Protection Ministry, about its policy plans. The results confirm that symbolic elements in communications increase citizens' trust in ill-perceived policies, by distracting them from scrutinizing their content. In contrast with my expectation, the effect of symbols was not attenuated by personal relevance. I discuss the implications of these findings for democratic responsiveness and accountability.*

@@

I further postulated, based on ELM, that the effect of symbols is attenuated the more the communication is personally relevant for citizens' lives. However, this latter expectation was not supported by the results. Residents of the polluted Haifa-Bay, who perceive EPM policy plan for reducing air-pollution in that area as highly relevant for their lives, were affected by the symbols not less than those who reside elsewhere. If anything, the results suggest that the symbols had a greater effect on Haifa-bay subjects. How can we reconcile these unexpected findings with the proposed theory? Given that the moderating role of personal relevance on peripheral cues was substantiated by numerous studies in social psychology, I would not argue that we should abandon this postulation based on this one study. One plausible explanation is that although Haifa-bay residents perceived the air-pollution policy as more relevant, their motivation to elaborate was not high enough to cause them to avoid peripheral cues. Certainly, the air-pollution in their area concerns them a great deal, as they point out themselves. However this issue represents a *continuous* concern that they may have got used to over the time, as opposed a new issue that emerged recently. If this explanation holds, this entails that citizens may resist the emotional effect of symbols only in rare circumstances, which are characterized by very high levels of personal relevance. These latter findings are also consistent with the findings of @alonbarkat_2017. In that study, the scholars assessed citizens' trust in the government electricity company, and its shaping by its familiar symbol, given citizens' recent personal experience of prolonged blackouts. Whilst their study was not purposely designed to test the interaction between symbols and personal relevance, one can assume that those who personally experienced blackouts perceived electricity service as more personally relevant. Still, they found that citizens who experienced blackouts were influenced by the symbol, as much as those who did not.

Certainly, we must be cautious when drawing general conclusions from our findings, for several particular reasons. First, this study focuses, empirically, on one public organization, which invested greatly in symbolic communications, most prominently, the use of high-profile celebrity comedians in its public campaigns. Future studies may further examine the proposed theory by focusing on additional cases of public organizations and policy areas. Moreover, in the symbols treatment group, I used a bundle of symbols, without disentangling their distinct effects. Future studies may compare the effectiveness of different types of symbolic elements. Finally, this study focuses on communication about policy plans, and therefore we must be cautious about generalizing from these findings to other types of communications. Specifically, future studies may seek to test my theory with regard to communications that include evidence about performance, and examine the interaction between symbols and information on good/bad outcomes.

Keeping these limitations in mind, this study has several important implications for public administration research. Most importantly, my findings highlight the significant role of branding and symbolic communication in shaping citizens' attitudes about government. Citizens' responses to symbolic elements in communications thus far received relatively little attention in Public administration literature, compared with the abounded research on their responses to information. Yet, this study, and others, demonstrated that symbols can be fairly effective in increasing citizens' trust, and mitigating their negative attitudes [@alonbarkat_2017; @karens_2016; @teodoro_2018]. Furthermore, my findings suggest that symbols decrease citizens' "rational", systematic processing of substantive information. This is good news as far as we are concerned mainly about citizens' *undervaluation* of government performance, and the negative biases that undermine the effectiveness of positive information.[^footnote_biases_ELM]. On the other hand, the emotive effect of government symbols on public opinion is highly problematic, insofar as it challenges the democratic ideal that citizens form their opinions about their government and its actions on the basis of rational thinking [@dahl_1973; @soroka_2010]. Moreover, it becomes especially problematic when symbols are used by poorly performing organizations,  or organizations that lack professional expertise. It those cases, symbolic elements can mitigate justified public criticism of government failures. Organizations can use them deliberately to distort public opinion via propaganda, or - more likely - without being aware of the tension between the symbols and their substantive message. Ultimately, by preserving citizens' trust in them and escaping from public blame, organizations can nurture their long-term positive reputation, which provides them political power and attenuates elected politicians' incentives to monitor and oversight them [@busuioc_2016; @bertelli_2018]. Hence, whereas government communication with citizens is normally perceived as a tool for enhancing democratic responsiveness and accountability, organizations can also use it as a means to undermine them.

@@

This study also contributes to existing behavioral Public Administration research by introducing ELM and applying it as a comprehensive theoretical framework for studying citizens' attitudes formation, and citizens' responses to government communications. ELM has various postulations and implications, but most importantly, it stresses that people respond differently to different communications, depending on people's motivation and ability to think about the subject, and on the quality of arguments and the peripheral cues in the communication. In this study, I applied ELM to distinguish between the peripheral processing of symbolic elements versus the central processing of substantive information, and to explore their moderation by citizens' motivation to elaborate on the message. Future studies may apply ELM and further explore the effects of other peripheral cues in government communications, and of other relevant motivation and ability factors, to enhance our understanding of the conditions under which government communications are likely to be more/less effective.

Finally, ELM has important methodological implications regarding the external validity of survey experiment studies of citizens' responses to government communications. The participants of these studies, including this particular one, have relatively high motivation to elaborate on the information that scholars presented them, because they may assume that they will be asked about it afterwards. In case they receive payment or other compensation for their participation, this further increases their motivation. By comparison, in the "real world", citizens are far less motivated to invest in thinking about government messages, and they probably also have lower ability to do so, given the high competition on their time and attention. Accordingly, their responses to performance information will less affected by asymmetrical processing of information, because most of them are not likely to process this information in the first place. Instead, most of them will rely on simple cues and heuristic shortcuts. I would expect, tentatively, that these differences exacerbate biases due to partisan and political ideology [@james_2017a; @baekgaard_2016] and implicit attitudes about the public-sector [@marvel_2015; @marvel_2015a; @hvidman_2016], while attenuating the role of negativity bias [@nielsen_2016; @olsen_2017]. Future studies may empirically explore these interactions. Regardless, scholars should also take to account this methodological issue when designing survey experiments, and attempt to simulate real world conditions, as much as possible. For instance, they can add filler sections in the survey before the real manipulation to attenuate participants' motivation and ability to elaborate. In addition, because the graphic design of communications also affects motivation and ability to elaborate, as shown here, scholars should simulate the design of real communications.
